# Reproducible Research: Peer Assessment 1
## Introduction
________________
This report will be discussing the data analysis about the activity monitoring dataset. The report will include the R code and the plots describing the data analysis. The structure of the report is as follows:
* Loading and preprocessing the data.
* Mean total number of steps taken per day and median total number of steps taken per day discarding the missing values.
* Average daily activity pattern.
* Imputing missing values and how imputing affects the data.
* Activity patterns during weekdays and weekends.

## Loading and preprocessing the data
The dataset is available in the github repository itself and hence reading the dataset directly from the repo.

```{r}
## Libraries of data.table and lattice are loaded to R
library(data.table)
library(lattice)
##Unzipping the file and reading it using read.csv()
unzippedfile <- unz("./activity.zip","activity.csv")
activityDF <- read.csv(unzippedfile, colClasses = c("numeric", "Date", "numeric"))
unlink(unzippedfile)
rm(unzippedfile)
## Converting the data frame to data table
activityDT <- data.table(activityDF)
## Number of columns is printed
ncol(activityDT)
## Number of rows is printed
nrow(activityDT)
head(activityDT,4)
tail(activityDT,4)
## Missing values NA are removed by using na.omit()
naremovedDT <- na.omit(activityDT)
head(naremovedDT,4)
tail(naremovedDT,4)
## number of rows without the missing values NA is printed
nrow(naremovedDT)
```

## What is mean total number of steps taken per day?
In this section, data analysis is preformed by calculating the mean total number of steps taken per day and the median total number of steps taken per day by discarding the missing values NA. The histogram of the total number of steps per day is plotted to examine the pattern.
```{r Plot1 Histogram }
## Total number of steps per day is calculated by using the function sum()
totalsteps <- naremovedDT[, sum(steps), by =date]
head(totalsteps,4)
hist(totalsteps$V1, col= "blue", xlab = "Total number of steps taken per day", main = "Histogram of Total number of steps taken each day", ylim = c(0,30))
meansteps <-mean(totalsteps$V1)
mediansteps <- median(totalsteps$V1)
format(meansteps, digits = 8)
format(mediansteps, digits = 8)
```
The mean total number of steps per day is `r format(meansteps, digits = 8)` and the median total number of steps per day is `r format(mediansteps, digits = 8)`. From the histogram, it is clear that the most of the total number of steps taken per day is between 10000 and 15000.

## What is the average daily activity pattern?
In this section, data analysis of which 5-minute interval has the maximum averaged steps taken across all days is examined. The plot is generated between the time series of 5-minute interval and the averaged steps across all the days. 
```{r Plot2 Lineplot}
## Averaged steps across all days is calculated by using the function mean()
averagesteps <- naremovedDT[, mean(steps), by=interval]
head(averagesteps,4)
tail(averagesteps,4)
str(averagesteps)
summary(averagesteps)
with(averagesteps, plot(interval,V1, type="l", col="purple", xlab = " 5 minute interval", ylab = "Averaged steps", main = "Averaged steps across all days Versus 5-minute interval"))
## 5- minute interval at which maximum averaged steps is calculated
## using which.max()
maxs <- which.max(averagesteps$V1)
maxin <- averagesteps$interval[maxs]
with(averagesteps, points(maxin,max(averagesteps$V1), pch = 1, col="blue"))
newvar <- paste0("( ", maxin)
newvaa <- paste0(newvar, " ,")
newvab <- paste0(format(max(averagesteps$V1), digits = 4), " )")
with(averagesteps, text(maxin, max(averagesteps$V1) + 5, paste0(newvaa,newvab), col= "blue"))
```

The maximum number of averaged steps across all the days - `r format(max(averagesteps$V1), digits = 4)` occurs at 5-minute interval of `r maxin`. There is broader region of least averaged steps across all the days and it occurs within the 5-minute interval of 0 and 500. Activity varies very significantly with the 5-minute interval.

## Imputing missing values
In this section, imputation technique is applied to the missing values by using the mean of all the steps (across all the days) for that 5- minute interval. Since thesteps averaged across all the days significantly varies with the 5-minute rather than with day, the steps averaged across all the days for that 5-minute interval is used to substitute the missing values NA in the original dataset.  
```{r}
## calculating the number of rows that has missing values NA for steps
naDT <- activityDT[is.na(activityDT$steps),][,]
nrow(naDT)
```
The number of rows of the original dataset that has missing values `r NA` for "steps" is `r nrow(naDT)`.

```{r Plot3 Histogram}
## calculating the mean of the steps grouped by interval and assigning to a ## new column in the data table
newactivityDT <- activityDT[,meansteps := mean(steps, na.rm= TRUE), by = interval]
## Re-assigning the calculated mean only to the missing values NA of the 
## column "steps""
newactivityDT$steps[which(is.na(newactivityDT$steps))] <- newactivityDT$meansteps[which(is.na(newactivityDT$steps))]
## New dataset created using imputation technique for missing values NA
nactivityDT <- newactivityDT[,!"meansteps", with=FALSE]
nrow(nactivityDT)
## Verifying that the new dataset contains NA
na.fail(nactivityDT)
## Calculating the total number of steps taken per day after imputation
naintotalsteps <- nactivityDT[, sum(steps), by =date]
head(naintotalsteps,4)
## Histogram is ploted
hist(naintotalsteps$V1, col= "green", xlab = "Total number of steps taken per day", main = "Histogram of Total number of steps taken each day", ylim = c(0,40))
## mean total number of steps per day is calculated
nainmeansteps <-mean(naintotalsteps$V1)
## median total number of steps per day is calculated
nainmediansteps <- median(naintotalsteps$V1)
format(nainmeansteps, digits = 8)
format(nainmediansteps, digits =8)
```
After imputation, it is evident from the histogram that the total number of steps taken per day has increased in frequency for the range of total number of steps - 10000 to 15000, as "mean"" is used for imputation. The mean continues to be the same value `r (format(nainmeansteps, digits = 8))` as mean is used to substitute the missing values. Median of the new data `r (format(nainmediansteps, digits =8))` is same as the value of the mean and has increased from the original and differs by `r mediansteps - nainmediansteps` from the median calculated discarding the missing values. For comparison, the histogram without missing values and histogram with imputation is plotted.
```{r Extraplot histograms}
## two historgrams are plotted side by side using function par()
par(mfrow = c(1,2))
hist(totalsteps$V1, col= "blue", xlab = "Total number of steps taken per day", main = "Histogram without NA's", ylim = c(0,40))
hist(naintotalsteps$V1, col= "green", xlab = "Total number of steps taken per day", main = "Histogram with imputation", ylim = c(0,40))
```

For comparison, the histogram without missing values and histogram with imputation is plotted above. 

## Are there differences in activity patterns between weekdays and weekends?
In this section, the activity patterns during weekdays and weekends are examined. 
```{r plot4 Latticeplot}
## new column of dayfactor is added 
newactivityDT[weekdays(date) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), dayfactor := as.factor("Weekday")]
newactivityDT[weekdays(date) %in% c("Saturday", "Sunday"), dayfactor := as.factor("Weekend")]
summary(newactivityDT)
str(newactivityDT)
## average of the steps across all the days are calculated after grouped
## by both 5-minute interval and dayfactor (weekday or weekend)
nainaveragesteps <- newactivityDT[, mean(steps), by="interval,dayfactor"]
head(nainaveragesteps,4)
tail(nainaveragesteps,4)
summary(nainaveragesteps)
str(nainaveragesteps)
## Lattice plot is generated
xyplot(nainaveragesteps$V1 ~ nainaveragesteps$interval | nainaveragesteps$dayfactor, layout = c(1,2), xlab = "5-minute Interval", ylab = "Averaged steps", panel = function(x,y, ...){
  panel.xyplot(x,y,type="l", col = "darkblue")
  })
```

From the lattice plot, it is evident that the activity pattern is oscillatory over broader range during weekend and there are no high sharp peaks. Whereas the activity pattern peaks up sharp and more during certain 5-minute interval (835) during weekdays. Also, it is clear that the maximum averagedsteps calculated in section "average daily activity pattern" occurs during weekdays.